Students' Alcohol Consumption Impacts on Academic Results on the Example of a Secondary Scool Maths Class in Portugal

### 2016-10-15 knit_MSproject.Rmd

```{r setup,include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@@

5. Interesting Findings

6. Conclusion

7. References

8. Appendices

@@

blank_theme <- theme_minimal()+ theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.border = element_blank(),
    panel.grid=element_blank(),
    axis.ticks = element_blank(),
    plot.title=element_text(size=14, face="bold")
  )
########################################################
# Sex
summary(sm[,'sex']) 
#F=208 #M=187
df_gender <- data.frame(Gender=c("Female", "Male"), value=c(208, 187))
pie_gender <- ggplot(data=df_gender, aes(x="", y=value, fill=Gender)) + 
  geom_bar(width = 1, stat = "identity") + 
  coord_polar("y", start=0) + 
  scale_fill_brewer() + blank_theme + 
  theme(axis.text.x=element_blank()) +
  geom_text(aes(y = value/3 + 
  c(0, cumsum(value)[-length(value)]), label = (value)), size=4, show.legend=FALSE) +
  ggtitle("Student's Gender")

###
# Address
summary(sm[,'address']) 
#Rural = 77 #Urban = 307
df_address <- data.frame(Type=c("Rural", "Urban"), value=c(77, 307))
pie_address <- ggplot(data=df_address, aes(x="", y=value, fill=Type)) + 
  geom_bar(width = 1, stat = "identity") + 
  coord_polar("y", start=0) + 
  scale_fill_brewer() + blank_theme + 
  theme(axis.text.x=element_blank()) +
  geom_text(aes(y = value/3 + c(0, cumsum(value)[-length(value)]), label = (value)), size=4) +
  ggtitle("Student's Home Address Type")

@@

#### a. General Information
General information about the students include gender, address and age. From the descriptive statistics (section 3) we can derive that the age from the students in this dataset ranges from 15 to 22 years with an average of 16.7 years. Out of the total of 395, there are 52.7% female and 47.3% male students. Most students would live in an urban environment, while approx. 20% live in an rural environment. 
```{r b, echo=FALSE}
grid.arrange(pie_gender, pie_address, ncol=2)
```
#### b. Family  
Information in the students families includes the family size, educational background of their parents and their jobs, and the student's guardian.

```{r c, echo=FALSE}
grid.arrange(pie_guardian, pie_fjob, pie_mjob, widths = c(1, 1, 2), layout_matrix = rbind(c(1,1,1), c(2,2,3)))
```

#### c. Education
The education section of this dataset covers the split between the two schools observed, why each student has chosen that particular school, and how long it takes each student to get to school. In addition, it provides information on the student's study time, extra educational support or classes taken and if the student is involved in extra curricular activities. Finally, there is data on their previous education and aspirations on whether to go into higher education.

```{r d, echo=FALSE}
grid.arrange(bar_studytime, bar_school, ncol=2)

@@

#### d. Social Life and Health 
The social life section provides information on activities the student is involved in outside of school. This includes how much leisure time they have, whether they are in a relationship and how often they go out. Moreover, there is data on how much alcohol students consume and in what health condition they are in.

#### e. Grades
The grades are split into three grades: the first period (G1), second period (G2) and final grade (G3). For the purpose of this analysis, students academic performance will be measured as the average of all three grades. In addition, there is information on whether they have failed to pass any classes.

```{r e, echo=FALSE}
pie_failures
```

@@

As aforementioned, this paper will look into whether alcohol is one of the factors affecting academic performance. Our basic assumption is alcohol consumption has a negative impact on students grade (Ho). To test this hypothesis, multivariate regression analysis will be applied on the 395 sample of students described above. This will allow us to see the significant factors that contribute to academic performance, and check whether or not alcohol consumption is one of them. In this regression, grade is assigned as the dependent variable, and all other variables are treated as independent variable. If alcohol appeared as one of the significant variables in the regression, the null hypothesis can not be rejected, inferring that alcohol **is** one of the variables related to academic performance.

@@

    If the distribution is highly skewed, linear regression might not give a good approximation or response varible.         Thus, variable transformation might be required, or applying other regression method might be more appropriate.

@@

    Correlation between response and explanatory variables is required to get a general view of the variables   relationships. In addition, correlation between explanatory variables will also be assessed to avoid multicollinearity problem, as well as to provide other insights from the data.

### a. Distribution of Response Variable

@@

ggplot(sm) + geom_density(aes(x=Score, fill = Score, colour = Score)) + labs(x = "Academic Score", y="% Frequency", title = "Academic Score of Secondary School Student")
```

As seen from the graph, the distribution is slightly left-skewed, but very close to normal distribution. Thus, multivariate regression method is appropriate to be applied to test the hypothesis, and no variable transformation required.

### b. Correlation Analysis

After performing correlation of all explanatory variables against grade (response variable), a few variables appear to have high correlation to grade :

1) Parent's Education
```{r g, echo=FALSE}
cor.test(sm[,"f_or_medu"],sm[,"Score"])
```
The correlation with grade is 0.23 With 95% confidence interval and p-value result of 3.586e-06, we can reject the null hypothesis of there is no correlation between these two variables. This means that parent's education has positive correlation with grade, and that students who have parents with university degree are more likely to have higher scores than those who do not.

2) Want to have higher education
```{r h, echo=FALSE}
cor.test(sm[,"higher_ed"],sm[,"Score"])
```
The correlation with grade is 0.189. With 95% confidence interval and p-value result of 0.0001517, we can reject the null hypothesis of there is no correlation between these two variables, and that grade does correlate positively with their aspiration to take higher education.

3) Number of past class failure
```{r i, echo=FALSE}
cor.test(sm$Score,sm$failures) 
```
The correlation with grade is -0.375. With 95% confidence interval and p-value result of 1.08e-14, we can reject the null hypothesis of there is no correlation between these two variables, and conclude that indeed, number of past class failure is negatively correlated with grade. students who have higher number of class failures in the past are more likely to have lower grade.

4) Frequency of going out with friends
```{r i1, echo=FALSE}
cor.test(sm$Score,sm$goout)
```

The correlation with grade is -0.154. With 95% confidence interval and p-value result of 0.002073, we can reject the null hypothesis of there is no correlation between these two variables, and conclude that indeed, the frequency of going out with friends are negatively correlated with grade. students who goes out more are more likely to have lower grade.

Alcohol does not have a significant correlation to grade, as shown in the correlation result result below :

```{r i2, echo=FALSE}
cor.test(sm$Score,sm$Walc)
```
The correlation shown is -0.08. With 95% confidence interval and p-value result of 8%, we can not reject the null hypothesis of there is no correlation between these two variables, and conclude that there is no correlation significant correlation between alcohol and students' academic performance.

suggestion: here we put remi's part. However, while performing correlation among explanatory variables .... (intro to remi's):

```{r i3, echo=FALSE}

###REMI's###

### Regression

After assessing the correlations between all the variables, multivariate regression will be applied. Firstly, all variables are put in the regression. Then, variable with the highest p-value is eliminated from the model before the next regression is performed. This iteration is done until a model with satisfying R2 and consists of significant explanatory variables is derived.

The end result of the regression is shown in the figure below.

Initial Regression :

```{r i4, echo=TRUE}
#Initial Regression

reg <- lm(Score ~ school + sex + age + address + famsize + Pstatus + Medu + Fedu + 
                Mjob + Fjob + reason + guardian + traveltime + studytime + failures + 
                schoolsup + famsup + paid + activities + nursery + higher + internet + 
                romantic + famrel + freetime + goout + Dalc + Walc + health + absences +
                f_or_mteach + f_or_medu
                , data = sm)
summary(reg)
```

@@

@@

#To cross check on the final model and the significant variables, we use existing R library:
#library(MASS)
#step <- stepAIC(reg, direction="both")
```

**Figure 4. Final Regression Model**

@@

library(stargazer)
stargazer(reg24, type = "text", digits = 2)
```

From the regression result, we can see that alcohol consumption is not one of the significant variables that explains student academic performance. Instead, they are:

@@

The R2 is 21.74%, which means that the model can explain 21.74% of the variance in student's academic score. The residual plot shown below (figure ...) does not show a random dispersed around the x axis, which means the residual does not sum up to zero. There are other significant factors not captured in the dataset nor the model, that can contribute in explaining academic score. Some examples are IQ score, EQ score, nutrition, or time spent for physical exercise.

**Figure 5. Residual Plot**

>>>>

@@

    + a. Distribution of Response Variable
    + b. Correlation Analysis
    + c. Regression

5. Conclusion

6. References

7. Appendices

@@

blank_theme <- theme_minimal()+ theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.border = element_blank(),
    panel.grid=element_blank(),
    axis.ticks = element_blank(),
    plot.title=element_text(size=14, face="bold")
  )
########################################################
# Sex
summary(sm[,'sex']) 
#F=208 #M=187
df_gender <- data.frame(Gender=c("Female", "Male"), value=c(208, 187))
pie_gender <- ggplot(data=df_gender, aes(x="", y=value, fill=Gender)) + 
  geom_bar(width = 1, stat = "identity") + 
  coord_polar("y", start=0) + 
  scale_fill_brewer() + blank_theme + 
  theme(axis.text.x=element_blank()) +
  geom_text(aes(y = value/3 + 
  c(0, cumsum(value)[-length(value)]), label = (value)), size=4, show.legend=FALSE) +
  ggtitle("Students' Gender")

###
# Address
summary(sm[,'address']) 
#Rural = 77 #Urban = 307
df_address <- data.frame(Type=c("Rural", "Urban"), value=c(77, 307))
pie_address <- ggplot(data=df_address, aes(x="", y=value, fill=Type)) + 
  geom_bar(width = 1, stat = "identity") + 
  coord_polar("y", start=0) + 
  scale_fill_brewer() + blank_theme + 
  theme(axis.text.x=element_blank()) +
  geom_text(aes(y = value/3 + c(0, cumsum(value)[-length(value)]), label = (value)), size=4) +
  ggtitle("Students' Home Address Type")

@@

**a. General Information**
General information about the students include gender, address and age. From the descriptive statistics (section 3) we can derive that the age from the students in this dataset ranges from 15 to 22 years with an average of 16.7 years. Out of the total of 395, there are 52.7% female and 47.3% male students. Most students would live in an urban environment, while approx. 20% live in an rural environment.

**Figure 1. General Information**
```{r b, echo=FALSE}
grid.arrange(pie_gender, pie_address, ncol=2)
```

**b. Family**  
Information in the students families includes the family size, educational background of their parents and their jobs, and the student's guardian.

**Figure 2. Family**
```{r c, echo=FALSE}
grid.arrange(pie_guardian, pie_fjob, pie_mjob, widths = c(1, 1, 2), layout_matrix = rbind(c(1,1,1), c(2,2,3)))
```

**c. Education**
The education section of this dataset covers the split between the two schools observed, why each student has chosen that particular school, and how long it takes each student to get to school. In addition, it provides information on the student's study time, extra educational support or classes taken and if the student is involved in extra curricular activities. Finally, there is data on their previous education and aspirations on whether to go into higher education.

**Figure 3. Education**
```{r d, echo=FALSE}
grid.arrange(bar_studytime, bar_school, ncol=2)

@@

**d. Social Life and Health**
The social life section provides information on activities the student is involved in outside of school. This includes how much leisure time they have, whether they are in a relationship and how often they go out. Moreover, there is data on how much alcohol students consume and in what health condition they are in.

**e. Grades**
The grades are split into three grades: the first period (G1), second period (G2) and final grade (G3). For the purpose of this analysis, students academic performance will be measured as the average of all three grades. In addition, there is information on whether they have failed to pass any classes.

**Figure 4. Past Class Failures**
```{r e, echo=FALSE}
pie_failures
```

@@

As aforementioned, this paper looks into factors influencing academic performance. Our expectation is to see a high correlation between family background and social activities on students' performances. In particular, we are interested to see how alcohol consumption is impacting students' grades. Our basic assumption is that alcohol consumption has a negative impact on students grade. Thus, the Null Hypothesis (H0) to be checked assumes that alcohol consumption does not have an impact on academic performance. To test this hypothesis, multivariate regression analysis will be applied on the 395 samples of students described above. This will allow us to see the significant factors that contribute to academic performance, and check whether or not alcohol consumption is one of them. In this regression, grade is assigned as the dependent variable, and all other variables are treated as independent variables. If alcohol appeared as one of the significant variables in the regression, the null hypothesis will be rejected, inferring that alcohol **is** one of the variables related to academic performance.

@@

If the distribution is highly skewed, linear regression might not give a good approximation of the response varible.         Thus, variable transformation might be required, or applying another regression method might be more appropriate.

@@

Correlation between response and explanatory variables is required to get a general view of the variables   relationships. In addition, correlation between explanatory variables will also be assessed to avoid a multicollinearity problem, as well as to provide other insights from the data.

#### a. Distribution of Response Variable

@@

ggplot(sm) + geom_density(aes(x=Score, fill = Score, colour = Score)) + labs(x = "Academic Score", y="% Frequency", title = "Academic Score of Secondary School Student")

summary(sm[,"Score"])
```

As seen from the graph, the distribution is slightly right-skewed, but very close to normal distribution. The deviation from the normal distribution could be due to a small sample size. Thus, applying a multivariate regression method is appropriate to test the hypothesis, and no variable transformation is required.

#### b. Correlation Analysis

After performing correlation of all explanatory variables against grade (response variable), four variables appear to have high a correlation to grade :

**1) Parent's Education**
```{r g, echo=FALSE}
cor.test(sm[,"f_or_medu"],sm[,"Score"])
```
The correlation between parent's education and grade is 0.23. With 95% confidence interval and a p-value of 3.586e-06, we can reject the null hypothesis, which states that there is no correlation between these two variables. This means that parent's education **has a positive correlation** with the student's grade, and that students who have parents with university degrees are more likely to have higher scores than those who do not.

**2) Want to have higher education**
```{r h, echo=FALSE}
cor.test(sm[,"higher_ed"],sm[,"Score"])
```
The correlation between students' aspirations to get higher education and grade is 0.189. With 95% confidence interval and a p-value of 0.0001517, we can reject the null hypothesis, which states that theres is no correlation between these two variables, and that grades do in fact correlate positively with their aspiration to take higher education.

**3) Number of past class failure**
```{r i, echo=FALSE}
cor.test(sm$Score,sm$failures) 
```
The correlation between number of past class failure and grade is -0.375. With 95% confidence interval and a p-value of 1.08e-14, we can reject the null hypothesis, which states that there is no correlation between these two variables, and conclude that indeed, the number of past class failures is negatively correlated with grades. Students who have failed more classes tend to have lower grades.

**4) Frequency of going out with friends**
```{r i1, echo=FALSE}
cor.test(sm$Score,sm$goout)
```

The correlation between the frequency of going out with friends and grade is -0.154. With 95% confidence interval and a p-value of 0.002073, we can reject the null hypothesis, which states that there is no correlation between these two variables, and conclude the frequency of going out with friends are negatively correlated with grades. Students who go out more often are more likely to have lower grades. In the following section, regression analysis will be performed to identify a fitting model which explains the students' academic performances and influencing factors.

**Correlation between Alcohol Consumption and Students' Performances**
```{r i2, echo=FALSE}
cor.test(sm$Score,sm$Walc)
```

As shown above, alcohol consumption does not have a significant correlation to students' grades. The test result shows a correlation of -0.08. With 95% confidence interval and a p-value of 8%, we can **not** reject the null hypothesis, which states that there is no correlation between these two variables. Therefore, we can conclude alcohol consumption does not have a significant impact on students' grades, which will also be verified in the following multivariate regression analysis.

#### c. Regression

After assessing the correlations between all the variables, multivariate regression will be applied. Firstly, all variables are put in the regression model. Secondly, the variable with the highest p-value is eliminated from the model. This iterative process is repeated until a model with a satisfying R2 value can be derived, which consists of significant explanatory variables.

Initial Regression Model :

```{r i4a, echo=TRUE}
#Initial Regression

reg <- lm(Score ~ school + sex + age + address + famsize + Pstatus + Medu + Fedu + 
                Mjob + Fjob + reason + guardian + traveltime + studytime + failures + 
                schoolsup + famsup + paid + activities + nursery + higher + internet + 
                romantic + famrel + freetime + goout + Dalc + Walc + health + absences +
                f_or_mteach + f_or_medu
                , data = sm)
```

@@

summary(reg)

@@

library(stargazer)
#To cross check on the final model and the significant variables, we use existing R library:
#library(MASS)
#step <- stepAIC(reg, direction="both")
```

**Figure 5. Final Regression Model**

The results of the iterative process are shown in the figure below.

@@

stargazer(reg24, type = "text", digits = 2)
```

These results verfify that alcohol consumption is not one of the significant variables that explains student academic performance. Instead, the dependent variables which best explain the model are following:

@@

The adjusted R2 is 21.74%, which means that the model can explain 21.74% of the variance in students' academic scores. The residual plot shown below (figure 6) does not show a randomly dispersed points around the x axis, which means that the residual does not sum up to zero. For predicted academic scores around 10, the model is closer to the observed values. For lower predicted scores, the model overestimates, and for higher scores, the predicted values are too low. There are other significant factors which are neither captured in the dataset nor the model, that can contribute to an explanation of students' academic scores. General knowledge indicates that IQ score, EQ score, nutrition, or time spent for physical exercise can have an impact on students' academic performance.

**Figure 6. Residual Plot**
